R Packages to Aid in Handling Web Access Logs

### 2015-12-26 paper/paper.Rmd
copyedit and style

---
title: R Packages to Aid in Handling Web Access Logs
author:
  - name: Oliver Keyes
    affiliation: Wikimedia Foundation
    address:
    - line 1
    - line 2
    email:  author1@work
  - name: Bob Rudis
    affiliation: Rapid7
    address:
    - line 1
    - line 2
    email:  author2@work
  - name: Jay Jacobs
    affiliation: Magical Python Land
    address:
    - line 1
    - line 2
    email:  author3@work
abstract: >
  Web access logs contain information on HTTP(S) requests and form a key part of both industry and academic explorations
  of human behaviour on the internet - explorations commonly performed in R. In this paper we explain and demonstrate
  a series of packages - \CRANpkg{webreadr}, \CRANpkg{urltools}, \CRANpkg{iptools} and \CRANpkg{rgeolocate} - designed to efficiently read in, parse
  and munge access log data, allowing researchers to handle it easily.
preamble: >
output: rticles::rjournal_article
---
## Introduction

@@

To make reading access logs into R as easy as possible we created the \CRANpkg{webreadr} package. This contains user-friendly equivalents to `read.table` for each type of log, detecting the fields that should appear, converting the timestamps into POSIX objects, and merging fields or splitting fields where necessary. The package contains four core functions, one for each form of log: `read_clf` for the Common Log Format, `read_combined` for the Combined Log Format, and `read_squid` and `read_aws` for Squid and AWS formats respectively. Each one abstracts away the complexity of specifying column names and formats, and instead allows a researcher to read a file in with a minimal amount of work.

@@

**webreadr** takes 1.2 seconds to read this file in at a minimum, and 1.48 at maximum, with a median of 1.33. This is consistently 3.5-6 times faster than the equivalent base R functionality - and is also, as explained, far easier to use.

@@

### Decoding
Both values can be percent-encoded, allowing them toinclude characters that are valid but reserved by the URL specification as having special meanings ("reserved characters"). A `#` symbol, for example, is encoded as <code>%23</code>: a percentage symbol, followed by the byte-code for that character.

@@

The solution is to be able to consistently decode URLs, which makes URL-based data far easier to analyse. Base R does contain a function, <code>URLdecode</code>, for doing just that, but is neither vectorised nor based on compiled code; with large datasets it takes an extremely long time.

To solve this common problem in analysing request logs, the \CRANpkg{urltools} [@urltools] package was created. This contains a function,
<code>url_decode</code>, which decodes URLs - but relies on vectorised, compiled code. Benchmarking the two approaches against each other shows that the \CRANpkg{urltools} implementation is approximately 60-70 times faster over large datasets. Again using **microbenchmark**, if we compare the vectorised decoding of 1,000,000 URLs with \CRANpkg{urltools} against `URLdecode` in a `vapply` loop, we see a 60-70 times speed improvement of urltools over base R:

\begin{figure}[h]
    \centering
    \includegraphics[scale=0.4]{decoding_benchmarks}
    \caption{Results of microbenchmark run: url\char`_decode versus base-R equivalent code}
\end{figure}

>>>>

---
title: R Packages to Aid in Handling Web Access Logs
author:
  - name: Oliver Keyes
    affiliation: Wikimedia Foundation
    address:
    - line 1
    - line 2
    email:  author1@work
  - name: Bob Rudis
    affiliation: Rapid7
    address:
    - line 1
    - line 2
    email:  author2@work
  - name: Jay Jacobs
    affiliation: Magical Python Land
    address:
    - line 1
    - line 2
    email:  author3@work
abstract: >
  Web access logs contain information on HTTP(S) requests and form a key part of both industry and academic explorations
  of human behaviour on the internet - explorations commonly performed in R. In this paper we explain and demonstrate
  a series of packages designed to efficiently read in, parse
  and munge access log data, allowing researchers to handle it easily.
preamble: >
output: rticles::rjournal_article
---
## Introduction

@@

To make reading access logs into R as easy as possible we created the \pkg{webreadr} package. This contains user-friendly equivalents to \code{read.table} for each type of log, detecting the fields that should appear, converting the timestamps into POSIX objects, and merging fields or splitting fields where necessary. The package contains four core functions, one for each form of log: \code{read\_clf} for the Common Log Format, \code{read\_combined} for the Combined Log Format, and \code{read\_squid} and \code{read\_aws} for Squid and AWS formats respectively. Each one abstracts away the complexity of specifying column names and formats, and instead allows a researcher to read a file in with a minimal amount of work.

@@

\pkg{webreadr} takes 1.2 seconds to read this file in at a minimum, and 1.48 at maximum, with a median of 1.33. This is consistently 3.5-6 times faster than the equivalent base R functionality - and is also, as explained, far easier to use.

@@

### Decoding
Both values can be percent-encoded, allowing them toinclude characters that are valid but reserved by the URL specification as having special meanings ("reserved characters"). A \samp{\#} symbol, for example, is encoded as \samp{\%23}: a percentage symbol, followed by the byte-code for that character.

@@

The solution is to be able to consistently decode URLs, which makes URL-based data far easier to analyse. Base R does contain a function, \code{URLdecode}, for doing just that, but is neither vectorised nor based on compiled code; with large datasets it takes an extremely long time.

To solve this common problem in analysing request logs, the \pkg{urltools} package was created. This contains a function,
\code{url\_decode}, which decodes URLs - but relies on vectorised, compiled code. Benchmarking the two approaches against each other shows that the \pkg{urltools} implementation is approximately 60-70 times faster over large datasets. Again using \pkg{microbenchmark}, if we compare the vectorised decoding of 1,000,000 URLs with \pkg{urltools} against \code{URLdecode} in a \code{vapply} loop, we see a 60-70 times speed improvement:

\begin{figure}[h]
    \centering
    \includegraphics[scale=0.4]{decoding_benchmarks}
    \caption{Results of microbenchmark run: \code{url\_decode} versus base-R equivalent code}
\end{figure}

### 2015-12-26 paper/paper.tex
copyedit and style

\abstract{
Web access logs contain information on HTTP(S) requests and form a key
part of both industry and academic explorations of human behaviour on
the internet - explorations commonly performed in R. In this paper we
explain and demonstrate a series of packages - \CRANpkg{webreadr},
\CRANpkg{urltools}, \CRANpkg{iptools} and \CRANpkg{rgeolocate} -
designed to efficiently read in, parse and munge access log data,
allowing researchers to handle it easily.
}

@@

To make reading access logs into R as easy as possible we created the
\CRANpkg{webreadr} package. This contains user-friendly equivalents to
\texttt{read.table} for each type of log, detecting the fields that
should appear, converting the timestamps into POSIX objects, and merging
fields or splitting fields where necessary. The package contains four
core functions, one for each form of log: \texttt{read\_clf} for the
Common Log Format, \texttt{read\_combined} for the Combined Log Format,
and \texttt{read\_squid} and \texttt{read\_aws} for Squid and AWS
formats respectively. Each one abstracts away the complexity of
specifying column names and formats, and instead allows a researcher to
read a file in with a minimal amount of work.

@@

\textbf{webreadr} takes 1.2 seconds to read this file in at a minimum,
and 1.48 at maximum, with a median of 1.33. This is consistently 3.5-6
times faster than the equivalent base R functionality - and is also, as
explained, far easier to use.

@@

Both values can be percent-encoded, allowing them toinclude characters
that are valid but reserved by the URL specification as having special
meanings (``reserved characters''). A \texttt{\#} symbol, for example,
is encoded as \%23: a percentage symbol, followed by the byte-code for
that character.

@@

The solution is to be able to consistently decode URLs, which makes
URL-based data far easier to analyse. Base R does contain a function,
URLdecode, for doing just that, but is neither vectorised nor based on
compiled code; with large datasets it takes an extremely long time.

To solve this common problem in analysing request logs, the
\CRANpkg{urltools} {[}@urltools{]} package was created. This contains a
function, url\_decode, which decodes URLs - but relies on vectorised,
compiled code. Benchmarking the two approaches against each other shows
that the \CRANpkg{urltools} implementation is approximately 60-70 times
faster over large datasets. Again using \textbf{microbenchmark}, if we
compare the vectorised decoding of 1,000,000 URLs with
\CRANpkg{urltools} against \texttt{URLdecode} in a \texttt{vapply} loop,
we see a 60-70 times speed improvement of urltools over base R:

\begin{figure}[h]
    \centering
    \includegraphics[scale=0.4]{decoding_benchmarks}
    \caption{Results of microbenchmark run: url\char`_decode versus base-R equivalent code}
\end{figure}

@@

\subsection{Summary}\label{summary}

This file is only a basic article template. For full details of
\emph{The R Journal} style and information on how to prepare your
article for submission, see the
\href{http://journal.r-project.org/latex/RJauthorguide.pdf}{Instructions
for Authors}.

@@

\address{
Oliver Keyes\\
Wikimedia Foundation\\
line 1\\ line 2\\
}
\href{mailto:author1@work}{\nolinkurl{author1@work}}

\address{
Bob Rudis\\
Rapid7\\
line 1\\ line 2\\
}
\href{mailto:author2@work}{\nolinkurl{author2@work}}

\address{
Jay Jacobs\\
Magical Python Land\\
line 1\\ line 2\\
}
\href{mailto:author3@work}{\nolinkurl{author3@work}}

>>>>

\abstract{%
Web access logs contain information on HTTP(S) requests and form a key
part of both industry and academic explorations of human behaviour on
the internet - explorations commonly performed in R. In this paper we
explain and demonstrate a series of packages designed to efficiently
read in, parse and munge access log data, allowing researchers to handle
it easily.
}

@@

To make reading access logs into R as easy as possible we created the
\pkg{webreadr} package. This contains user-friendly equivalents to
\code{read.table} for each type of log, detecting the fields that should
appear, converting the timestamps into POSIX objects, and merging fields
or splitting fields where necessary. The package contains four core
functions, one for each form of log: \code{read\_clf} for the Common Log
Format, \code{read\_combined} for the Combined Log Format, and
\code{read\_squid} and \code{read\_aws} for Squid and AWS formats
respectively. Each one abstracts away the complexity of specifying
column names and formats, and instead allows a researcher to read a file
in with a minimal amount of work.

@@

\pkg{webreadr} takes 1.2 seconds to read this file in at a minimum, and
1.48 at maximum, with a median of 1.33. This is consistently 3.5-6 times
faster than the equivalent base R functionality - and is also, as
explained, far easier to use.

@@

Both values can be percent-encoded, allowing them toinclude characters
that are valid but reserved by the URL specification as having special
meanings (``reserved characters''). A \samp{\#} symbol, for example, is
encoded as \samp{\%23}: a percentage symbol, followed by the byte-code
for that character.

@@

The solution is to be able to consistently decode URLs, which makes
URL-based data far easier to analyse. Base R does contain a function,
\code{URLdecode}, for doing just that, but is neither vectorised nor
based on compiled code; with large datasets it takes an extremely long
time.

To solve this common problem in analysing request logs, the
\pkg{urltools} package was created. This contains a function,
\code{url\_decode}, which decodes URLs - but relies on vectorised,
compiled code. Benchmarking the two approaches against each other shows
that the \pkg{urltools} implementation is approximately 60-70 times
faster over large datasets. Again using \pkg{microbenchmark}, if we
compare the vectorised decoding of 1,000,000 URLs with \pkg{urltools}
against \code{URLdecode} in a \code{vapply} loop, we see a 60-70 times
speed improvement:

\begin{figure}[h]
    \centering
    \includegraphics[scale=0.4]{decoding_benchmarks}
    \caption{Results of microbenchmark run: \code{url\_decode} versus base-R equivalent code}
\end{figure}

@@

Direct speed comparisons aren't possible, since the functionality it
provides is not (to the authors' knowledge) replicated in other R
packages, but it is certainly faster (and more secure) than
internet-dependent alternatives.

\subsection{Conclusions and further
work}\label{conclusions-and-further-work}

In this research article we have demonstrated a set of tools for
handling access logs during every stage of the data cleaning pipeline -
reading them in with \pkg{webreadr}, decoding, manipulating and
extracting value from URLs with \pkg{urltools}, and retrieving
geographic metadata from IP addresses with \pkg{iptools} and
\pkg{rgeolocate}. We have also demonstrated the dramatic speed
improvements in using these tools in preference to existing methods
within R.

Further work - integrating more sources of geolocation information
within \pkg{rgeolocate}, supporting UTF-8 URL decoding within
\pkg{urltools}, and increasing IPv6 support in \pkg{iptools} - would
make these packages even more useful to Human-Computer Interaction
researchers and other specialists who rely on access logs as primary
data sources. Even without that functionality, however, this suite of
packages is a dramatic improvement upon the status quo.

@@

\address{%
Oliver Keyes\\
Wikimedia Foundation\\
line 1\\ line 2\\
}
\href{mailto:author1@work}{\nolinkurl{author1@work}}

\address{%
Bob Rudis\\
Rapid7\\
line 1\\ line 2\\
}
\href{mailto:author2@work}{\nolinkurl{author2@work}}

\address{%
Jay Jacobs\\
Magical Python Land\\
line 1\\ line 2\\
}
\href{mailto:author3@work}{\nolinkurl{author3@work}}
